Region-network hierarchical sparsity priors for\\ high-dimensional inference in brain imaging

### 2016-01-27 paper/paper.tex
final touches in intro+disc

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., ``long data")  to
the high-dimensional regime (i.e., ``wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This high-dimensional scenario challenges many statistical estimators from
classical statistics.
For instance,
the estimation of any generalized linear model without additional assumptions
becomes underdetermined, i.e. yields full subspace of possible solutions, many or most of which will not generalize well in a predictive setting.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{buhlmann2011statistics, hastie2015statistical}, which act as a 
regularizer and can be used for model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Generally, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus preassuming the variables to have unequal importance
and to obey expected data distributions
\cite{bach2012optimization}.
{\color{red}Michael: This is the definition of a prior, not of strutured sparsity}

%% Yet, what neurobiological structure suggests itself
%% to harness the
%% \textit{curse of dimensionality} with
%% $\textgreater$100,000 variables
%% in neuroimaging research?
In neuroimaging research, generalized linear models predicting external 
variables from brain images, which can reach a dimensionality of several
hundreds of thousands of voxels, necessarily need appropriate regularization.
Which neurobiological prior knowledge lends itself to the construction of a
structured sparse prior?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of anatomy (i.e., histological preparation),
connectivity, (i.e., axonal tracing),
and functional properties
(i.e., single cell recordings) in animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local
%cyto- and chemoarchitectonic
infrastructure
and its unique global connectivity profile
together are thought to realize brain function.
%
In sum,
cortical brain modules versus connections between them
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation}.
Importantly,
probably no existing brain analysis method acknowledges that
both architectural principles are inextricably involved
in the realization of mental operations
\cite{tononi1998complexity, saygin2012}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
for instance,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
was later enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
Today,
high-thoughput approaches enable
ultrahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
%
Invasive axonal tracing studies in monkeys were complemented
by diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
These revealed the important implication of
canonical brain networks across cognitive domains,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these macroscopical networks
were repeatedly observed to be induced
by the onset of given cognitive tasks \cite{fransson2006}.
Such task-induced mechanisms orchestrating supraordinate networks
might be subserved by the right anterior insula \cite{sridh2008}
and temporo-parietal junction \cite{bzdok2013tpj}.
%
Ultimately,
interpretation of findings from all these methods naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these two major interpretational streams in systems neuroscience,
the present study proposes to incorporate
established neurobiological structure underlying
functional segregation and integration
into supervised estimators
by hierarchical structured sparsity.
% The "true" relative importance of 
% local region compartments and global network compartments
% is typically unknown
% but
% probably varying in degree
% across diverse neuroscientific questions.
%
Learning techniques
exploiting structured sparsity 
have recently made much progress in various application domains
from processing of auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured, kim2010sparse}
to
genetics \cite{rapaport2008classification, kim2012tree},
astrophysics \cite{vinci2014estimating},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.
%
This is extended by the present work that
enables neuroscience-specific estimators capitalizing on
neurobiologically plausible region and network priors.
%
Using a large reference dataset,
we demonstrated that domain-informed supervised models
gracefully tackle the curse of dimensionality,
yield more human-interpretable results,
and generalize better to new samples
than domain-na\"ive estimators.

@@

\section{Discussion}
Relevant structure in brain imaging data has long been investigated
separately
according to two distinct organizational principles:
functional segregation into discrete brain regions
and functional integration by interregional brain networks.
%
This paper demonstrates the simultaneous exploitation of
both these neurobiological compartments
for sparse variable selection and high-dimensional prediction
in a reference dataset.
%
Introducing existing domain knowledge into model selection
allowed privileging members of the function space
that are most neurobiologically plausible.
%
Domain-informed hierarchical structured sparsity is shown to enhance
both model interpretability and generalization performance,
although these statistical-learning goals are typically in conflict.

The present approach has important advantages over previous
neuroimaging studies that capitalized on dimensionality reduction to harness
the curse of dimensionality.
They often used preliminary region-wise pooling functions
or regression against network templates
for subsequent supervised learning on the aggregated feature space.
Such two-step approaches of feature engineering and inference
\textit{i)} can only account for either functional specialization or
functional integration of brain organization,
\textit{ii)} depend on the ground truth being a region or network effect,
and
\textit{iii)} cannot issue individual coefficients for every brain voxels.
%
Hierarchical region-network sparsity overcomes these shortcomings
by estimating individual voxel contributions
while benefitting from their functional segregation and integration
to restrict statistical complexity.
%
Viewed from the bias-variance tradeoff,
our modification to logistic regression estimators
entailed a large decrease in model variance but only a modest
increase in model bias.
Viewed from the Vapnik-Chervonenkis dimensions,
this entailed a healthy decrease in the complexity capactity of the prediction model
with a higher chance of generalizing to unobserved data.

>>>>

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., ``long data")  to
the high-dimensional regime (i.e., ``wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This high-dimensional scenario challenges various statistical methods from
classical statistics.
For instance,
estimating generalized linear models without additional assumptions
yields an underdetermined system of equations.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{buhlmann2011statistics, hastie2015statistical}.
They act as a 
regularizer and can be used for model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Generally, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus shrinking and selecting variables guided by
expected data distributions
\cite{bach2012optimization}.
Such restrictions to complexity are an attractive plan of attacked
for the $\textgreater$100,000 variables of brain maps.
Yet, what generally accepted neurobiological structure lends itself
to harness the
\textit{curse of dimensionality} by structured sparsity priors?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of anatomy (i.e., histological preparation),
connectivity (i.e., axonal tracing),
and functional properties
(i.e., single-cell recordings) in animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local
%cyto- and chemoarchitectonic
infrastructure
and its unique global connectivity profile
together are thought to realize brain function.
%
In sum,
cortical brain modules versus connections between them
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation}.
Importantly,
probably no existing brain analysis method acknowledges that
both functional organizations are inextricably involved
in the realization of mental operations
\cite{tononi1998complexity, saygin2012}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single-cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
was later enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
High-thoughput approaches today enable
ultrahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
%
Invasive axonal tracing studies in monkeys were complemented
by diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
These revealed the important implication of
canonical brain networks across psychological tasks,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these macroscopical networks
were repeatedly observed to be induced
by the onset of given psychological tasks \cite{fransson2006}.
% Such task-induced mechanisms orchestrating supraordinate networks
% might be subserved by the right anterior insula \cite{sridh2008}
% and temporo-parietal junction \cite{bzdok2013tpj}.
%
As a common point of all these methods,
interpretation of findings naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these two major interpretational streams in systems neuroscience,
the present study proposes to incorporate
established neurobiological structure underlying
functional segregation and integration
into supervised estimators
by hierarchical structured sparsity.
% The "true" relative importance of 
% local region compartments and global network compartments
% is typically unknown
% but
% probably varying in degree
% across diverse neuroscientific questions.
%
Learning techniques
exploiting structured sparsity 
have recently made much progress in various application domains
from processing of auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured, kim2010sparse}
to
astrophysics \cite{vinci2014estimating},
genetics \cite{rapaport2008classification, kim2012tree},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.
%
This is extended by the present work that
introduced neuroscience-specific estimators capitalizing on
neurobiologically plausible region and network priors.
%
Based on the largest neuroimaging repository,
we demonstrated that domain-informed supervised models
gracefully tackle the curse of dimensionality,
yield more human-interpretable results,
and generalize better to new samples
than domain-na\"ive estimators.

@@

\section{Discussion}
Relevant structure in neuroimaging data has long been investigated
according to two separate organizational principles:
functional segregation into discrete brain regions
and functional integration by interregional brain networks.
%
This paper demonstrates the simultaneous exploitation of
both these neurobiological compartments
for sparse variable selection and high-dimensional prediction
in a reference dataset.
%
Introducing existing domain knowledge into model selection
allowed privileging members of the function space
that are most neurobiologically plausible.
%
Domain-informed hierarchical structured sparsity is shown to enhance
both model interpretability and generalization performance,
although these statistical-learning goals are typically in conflict.

The present approach has important advantages over previous
neuroimaging studies that capitalized on dimensionality reduction to harness
the curse of dimensionality.
They often used preliminary region-wise pooling functions
or regression against network templates
for subsequent supervised learning on the aggregated feature space.
Such lossy two-step approaches of feature engineering and inference
\textit{i)} can only account for either functional specialization or
functional integration of brain organization,
\textit{ii)} depend on the ground truth being a region or network effect,
and
\textit{iii)} cannot issue individual coefficients for every brain voxels.
%
Hierarchical region-network sparsity overcomes these shortcomings
by estimating individual voxel contributions
while benefitting from their functional segregation and integration
to restrict statistical complexity.
%
Viewed from the bias-variance tradeoff,
our modification to logistic regression estimators
entailed a large decrease in model variance but only a modest
increase in model bias.
Viewed from the Vapnik-Chervonenkis dimensions,
this entailed a healthy decrease in the complexity capacity of the prediction model
with a higher chance of generalizing to unobserved data.
